---
title: "Bike Rental Report"
author: "Louise Braithwaite"
date: "31/07/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=normalizePath('~/Bike_rental_project/Bike_Rental_Report'))
```

``` {r loading, include=FALSE}
library(ProjectTemplate)
load.project()
```

## 1. Business Overview

Over a two year period a bike rental company has collected data on the number of daily rentals, daily weather conditions (temperature, humidity, wind speed and precipitation type) and day type (season, year, month, holiday, weekday, working day). The company has commissioned the NICD to analyse and deliver business insights that will help them understand how the change in weather effects the number of bikes they rent each day.

### 1.1 Data mining goals and success criteria

This project will focus on data description and summarisation of the data set to help develop insights into the impact weather has on bike rentals.

The goal of this project is to:

__*Identify how the change in weather effects the number of bikes rented each day.*__


### 1.2 Requirements, assumptions, and constraints

Requirements:

* A report with a maximum of 500 words and 2 figures, detailing how the data has been interpreted and what insights have been gained.
* A Git repository of the work completed
* A ProjectTemplate folder containing reports and source code

Assumptions and constraints: 

* We can only work with the data set provided
* No meta date has been provided. All data assumptions should be logged so they can be verified later


## 2. Data Overview and Preparation
A data set was provided by the bike rental company as a csv file. The data set is complete with no missing values. 

The next step was to investigate what information is contained within the data set and evaluate which variables relate to the project goal. The section below contain:

* a snapshot of the first 5 observations of the data file. This provides the column headers and vector class
* a summary table of the variables in the provided data set
* a list of the key variables associated with the project goal
* a summary of data assumptions
* data preparation notes
* table summarising the final two data frames
* a summary table of the variables in the final data set



## 2.1 Data Overview

Review the variables in the original bike.rental.data.csv file. In R the original data (bike.rental.data.csv) is saved as a data frame named 'rental.data'.

``` {r initial review: rental.data} 
head(rental.data, 5)
```
.

_**Table 1: Summary of the variables from bike_rental_data csv file**_ 

Column Header | Class | Example | Description
------------- | ------------- | ------------- | -------------
season  | character | "SPRING" | The season, spring, summer, fall or winter
yr  | integer | 2011 | The year, either 2011 or 2012
mnth  | character | "JAN" | The month, all in shortened character form (e.g. "JAN", "FEB", "MAR")
holiday | character | "NO HOLIDAY"  | Whether the day is a holiday "HOLIDAY" or not "NO HOLIDAY"
weekday | character | "SAT" | The day of the week, all in shortened character form (e.g. "SAT", "SUN", "MON")
workingday | character | "NO WORKING DAY"  | Whether it is a "WORKING DAY" or a "NO WORKING DAY"
weathersit | character | "MISTY" | Weather category ("GOOD", "MISTY" or "RAIN/SNOW/STORM")
temp  | numeric | 8.18 | Temperature (degrees celsius, °C) 
hum  | numeric | 80.6 | Relative humidity (%)
windspeed  | numeric | 10.7 | Wind speed (mph)
cnt  | integer | 985 | The number of bikes rented that day (rental count) 
days_since_2011  | integer | 0,1,2 | A sequence of numbers starting a 0 on 1 January 2011 and increasing by 1 each day

#### 2.1.1 Key Variables

There are five variables we can use to help us determine how weather can effect the number of bicycle rentals. They are:

* cnt
* weathersit
* temp
* hum
* windspeed

#### 2.1.1 Data Assumptions  

The following data assumptions were made. They should be verified with the data owner at the next opportunity. 

* Temperature is measured in degrees celsius (°C) 
* Humidity measure is relative humidity (%)
* Wind speed is measured in miles per hour (mph)
* Temperature, humidity and wind speed can be rounded to integers, as this level of detail is most appropriate and will make analysis and understanding easier. 
* Weathersit variables are an indication of precipitation, "GOOD" can also be referred to as "NO.RAIN".

### 2.2 Data Preparation Notes  
The following data preparation tasks were performed to ease the exploratory data analysis process.

1. To ease navigate through the data, create proper date variables to replace 'yr', 'month', 'weekday' and 'days_since_2011'. 
i) Create date column, use the 'days_since_2011' variable to quality check the new date variable. Eg. observation 31 of 'days_since_2011' is 30 but this equates to 31 January 2011
ii) Introduce the lubridate package to create new date variables, 'just.day', 'just.dayofweek', 'just.dayofweek2' (character labels), 'just.month', 'just.month2' (character labels) and 'year'
iii) Remove the original date columns, which are no longer needed ('yr', 'mnth', 'weekday')
2. Convert categorical character variables to factors ('season', 'holiday', 'workingday' and 'weathersit')
3. Rename variables for clarity and to create naming convention ('cnt' to 'rental.count', 'weathersit' to 'weather.category', 'temp' to 'temperature, 'hum' to 'humidity', 'windspeed' to 'wind.speed', 'workingday' to 'working.day' and 'days_since_2011' to 'days.since.2011')
4. Round the 'temperature', 'wind.speed' and 'humidity' columns so they become integers
5. Create new column called 'precipitation' using the data from 'weather.categories' and change the characters to numbers to reflect the factor levels
6. Reorder variables so the response variable is listed first

.  
.  
.  
.  
.  
.  
.  
.  
.  
.  
.  
.  
.  
.  
.  
.  




#### 2.2.1 Updated Data Frame  
.  


_**Table 2: Summary of the variables from data frame after data has been prepared**_ 

Column Header | Class | Example | Description
------------- | ------------- | ------------- | -------------
rental.count  | integer | 985 | The number of bikes rented that day 
precipitation | numeric | 1 | Weather category (1 = "GOOD", 2 = "MISTY" and 3 = "RAIN/SNOW/STORM")
temperature  | numeric | 8.18 | Temperature (degrees celsius, °C) 
humidity  | numeric | 80.6 | Relative humidity (%)
wind.speed  | numeric | 10.7 | Wind speed (mph)
weather.category | character | "MISTY" | Weather category ("GOOD", "MISTY" or "RAIN/SNOW/STORM")
season  | character | "SPRING" | The season, spring, summer, fall or winter
holiday | character | "NO HOLIDAY"  | Whether the day is a holiday "HOLIDAY" or not "NO HOLIDAY"
working.day | character | "NO WORKING DAY"  | Whether it is a "WORKING DAY" or a "NO WORKING DAY"
days.since.2011  | integer | 0,1,2 | A sequence of numbers starting a 0 on 1 January 2011 and increasing by 1 each day
date  | date | "2011-01-01" | The date
just.day  | integer | 1 | The day of the month
just.dayofweek  | numeric | 1 | The day, 1-7 (1 = Sunday, 2 = Monday, etc)
just.dayofweek2  | factor | "Sun" | The day, all in shortened character form (e.g. "Sun", "Mon", "Tue")
just.month | numeric | "JAN" | The month, 1-12 (1 = Jan, 2 = Feb etc)
just.month2  | factor | "Jan" | The month, all in shortened character form (e.g. "Jan", "Feb", "Mar")
year  | numeric | 2011 | The year, either 2011 or 2012

.  
.  
.  


#### 2.2.2 Final Data Frames  


After preparing the data we are left with two data frames.


_**Table 3: Summary of the two final data frames**_   

Data Frame | Description | Dimensions
------------- | ------------- | -------------
rental.data  | The original data  | 731 obs. and 12 variables
bike.rental.data  | Data preparation steps 1-6 applied | 731 obs. and 17 variables

### 3. Exploratory Data Analysis  

#### Pairs Plot   

Create a pairs plot of the variables relating to rental count and weather.

```{r pairs plot, echo=FALSE}
# subset key variables to create new data frame
bike.rental.key = bike.rental.data[,1:5]
# Produce a scatterplot matrix of predictor and explanatory variables
pairs(bike.rental.key)

```

Findings from pairs plot:

* There appears to be a moderately strong and positive linear correlation between rental.count and temperature; * as temperature increases so does rental.count.
* There appears to be some indication of a negative linear correlation between rental.count and humidity; as humidity increases rental.count decreases.
* There appears to be some indication of a negative linear correlation between rental.count and wind speed; as wind speed increases rental.count decreases.
* The pairs plot indicates that the weather.category is an indicator of precipiation, as the "MISTY" and "RAIN/SNOW/STORM" categories have higher recording of humidity. *It seems safe to assume that good indicates no rain.* 


```{r precipitation, message = FALSE}
# Review the categorical precipitation data
table(bike.rental.key$precipitation)
```

From the summary above table we can see that weather has been categorised as "GOOD" 63% of the time. 34% of the time the weather is categorised as "MISTY" and 3% of the time the weather is categorised as "RAIN/SNOW/STORM".



```{r sample means and covariance matrix, echo=FALSE, include = FALSE}
#### Sample Means and Covariance Matrix
# calculate the variable means
colMeans(bike.rental.key)
# calculate the covariance matrix
var(bike.rental.key)
```


#### Correlation Matrix  

Use the correlation matrix to quantify the strength of a linear relationship between pairs of variables. 

```{r correlation matrix, message=FALSE}
# calculate the correlation matrix
cor(bike.rental.key)
```

The correlation matrix confirms some of our previous observations:

* The correlation between rental.count and temperature, 0.63, indicates a moderately strong positive linear relationship
* The correlation between rental.count and weather.catergory, -0.3, suggests there is some indication of a negative linear relationship between the two variables. This supports the assumption that as the weather goes from no rain ("GOOD"), to "MISTY", to "RAIN/SNOW/STORM" the rental.count will decrease. 
* The correlation between weather.category and humidity, 0.59, indicates a moderately strong positive linear relationship. Observation makes sense, as we would expect the air to be more humid if the weather is misty or it is precipitating.  
* The correlation between rental.count and wind.speed, -0.24, and rental.count and humidity, -0.1, are negative but are too low to indicate a linear relationship

#### Correlation plots  

The correlation plots are ordered to reflect the weather variables' degree of correlation. Temperature has the greatest linear correlation to rental count, then precipitation, wind speed and finally humidity. 

```{r correlation summary, echo=FALSE, message = FALSE}
# create plot showing correlation between rental.count and precipitation, temperature, humidity and wind.speed
library(cowplot)
library(gridExtra)
library(ggpubr)

### Fit Least Square Regression Line ####
lsq_fit_temperature = lm(rental.count ~ temperature, data = bike.rental.key)
lsq_fit_precipitation = lm(rental.count ~ precipitation, data = bike.rental.key)
lsq_fit_wind.speed = lm(rental.count ~ wind.speed, data = bike.rental.key)
lsq_fit_humidity = lm(rental.count ~ humidity, data = bike.rental.key)

plotA = ggplot(bike.rental.key) +
  geom_point(aes(x=temperature, y=rental.count), color = "#02075d", shape=4) +
  geom_abline(intercept = coefficients(lsq_fit_temperature)[1], 
              slope = coefficients(lsq_fit_temperature)[2], 
              col="red", linetype = "dashed", size = 1) +
  ggtitle(label = "Daily rental count by temperature") + 
  labs(x="Temperature (°C)", y="Daily rental count") +
  theme_classic() +
  theme(plot.subtitle = element_text(color = "red", size = 8),
        plot.title = element_text(size = 10),
        axis.line = element_line(colour = "#a9a9a9"))
               

plotB = ggplot(bike.rental.key) +
  geom_point(aes(x=precipitation, y=rental.count), color = "#02075d", shape=18) +
  geom_abline(intercept = coefficients(lsq_fit_precipitation)[1], 
              slope = coefficients(lsq_fit_precipitation)[2], 
              col="red", linetype = "dashed", size = 1) +
  ggtitle(label = "Daily rental count by precipitation") + 
  labs(x="Precipitation category", y="Daily rental count") +
  theme_classic() +
  theme(plot.subtitle = element_text(color = "red", size = 8),
        plot.title = element_text(size = 10),
        axis.line = element_line(colour = "#a9a9a9"))               

plotC = ggplot(bike.rental.key) +
  geom_point(aes(x=wind.speed, y=rental.count), color = "#02075d", shape=4) +
  geom_abline(intercept = coefficients(lsq_fit_wind.speed)[1], 
              slope = coefficients(lsq_fit_wind.speed)[2], 
              col="red", linetype = "dashed", size = 1) +
  ggtitle(label = "Daily rental count by wind speed") + 
  labs(x="Wind speed (mph)", y="Daily rental count") +
  theme_classic() +
  theme(plot.subtitle = element_text(color = "red", size = 8),
        plot.title = element_text(size = 10),
        axis.line = element_line(colour = "#a9a9a9"))   

plotD = ggplot(bike.rental.key) +
  geom_point(aes(x=humidity, y=rental.count), color = "#02075d", shape=4) +
  geom_abline(intercept = coefficients(lsq_fit_humidity)[1], 
              slope = coefficients(lsq_fit_humidity)[2], 
              col="red", linetype = "dashed", size = 1) +
  ggtitle(label = "Daily rental count by humidity") + 
  labs(x="Realtive humidity (%)", y="Daily rental count") +
  theme_classic() +
  theme(plot.subtitle = element_text(color = "red", size = 8),
        plot.title = element_text(size = 10),
        axis.line = element_line(colour = "#a9a9a9"))   

figure = plot_grid(plotA, plotB, plotC, plotD,
          labels = c("A", "B", "C", "D"),
          ncol = 2,
          nrow = 2)

annotate_figure(figure,
                top = text_grob("Correlation plots of daily rental count against weather variables", 
                                size = 11, hjust = 0.5, x =0.5),
                                bottom = text_grob("Red dashed line = least square regression line", color = "red", size =10, hjust = 0, x =0),
                )
```
#### Summary  

The key finding from this analysis is that temperature has the most significant linear relationship with rental count. We will explore this relationship further in the next section.


#### Temperature   

The mean temperature across the two years is 15 degrees, however the histogram shows that the distribution of the temperature data is bimodal. The first peak is at 12 degrees and the other at around 31 degrees. 

``` {r temperature, echo = FALSE}

ggplot(data = bike.rental.data) +
  geom_histogram(mapping = aes(x=temperature), binwidth = 3) +
  scale_x_continuous(breaks = round(seq(min(bike.rental.data$temperature), max(bike.rental.data$temperature), by = 5),1)) +
  theme_classic()+
  ggtitle(label = "Temperature histogram")


# The data is roughly bimodal. The first peak is at 12 degrees and the other at around 31 degrees

summary(bike.rental.data$temperature)
```

Let's plot temperature and rental count across the year onto one plot (figure 1).

``` {r rental count and temperature by day, echo=FALSE}
# Plot temperature and rental count over time with a dual Y axis
# value used to transform the data

coeff = 100

# Select colours
temperatureColour = "#FFA500"
rentalColour = "#02075d"

# plot
ggplot(bike.rental.data, aes(x=date)) +
  
  geom_point(aes(y=temperature), color = temperatureColour) +
  geom_point(aes(y=rental.count/ coeff), color = rentalColour) +
  
  scale_y_continuous(
    # Features of the first axis
    name = "Temperature (°C)",
    breaks = round(seq(min(bike.rental.data$temperature), max(bike.rental.data$temperature), by = 5),1),
    #Add a second axis and specify its features
    sec.axis = sec_axis(~.*coeff, name = "Number of rentals per day")
  ) +
  
  theme_classic() +
  theme(axis.title.y = element_text(color = temperatureColour, size=12, vjust = 1),
        axis.title.y.right = element_text(color = rentalColour, size=12, vjust = 1),
        axis.title.x = element_text(size=12),
        plot.subtitle = element_text(face = "bold", size = 12),
        plot.title = element_text(face = "bold", size = 12),
        plot.caption = element_text(size = 12)) +
  ggtitle(label = "Figure 1",
          subtitle = "Rental count and temperature per day") +
  labs(caption = "Data source: bike.rental.data") + 
  xlab("Date")



```

The daily rental count follows the temperature pattern across the two years.

``` {r average rentals by temperature, echo=FALSE, message = FALSE}
rentalsBYtemp = bike.rental.data %>% 
  group_by(temperature, year) %>%
  summarise(totalrentals = sum(rental.count),
            median.daily.rentals = median(rental.count),
            mean.daily.rentals = mean(rental.count),
            sd.daily.rentals = sd(rental.count),
            percentage = round((totalrentals/3292679)*100)) # 3,292,679 = sum(bike.rental.data$rental.count))

rentalsBYtemp$year = as.character(rentalsBYtemp$year)

highlight_df <- rentalsBYtemp %>% 
  group_by(year) %>%
  filter(mean.daily.rentals == max(mean.daily.rentals))

ggplot(rentalsBYtemp) +
  
  geom_point(aes(x=temperature, y=median.daily.rentals, color = year)) +
  geom_line(data = filter(rentalsBYtemp, year == 2012), aes(x=temperature, y=median.daily.rentals), color = "#02075d") +
  geom_line(data = filter(rentalsBYtemp, year == 2011),  aes(x=temperature, y=median.daily.rentals), color = "#FFA500") +
  geom_point(data=highlight_df, aes(x=temperature,y=median.daily.rentals), color='red') +
  geom_text(data=highlight_df, aes(x=temperature,y=median.daily.rentals, label= as.character(temperature)),
            hjust=1,vjust=0, nudge_x = 2, colour = "red", fontface = "bold") +
  
  theme_classic() +
  theme(plot.subtitle = element_text(face = "bold", size = 12),
        plot.title = element_text(face = "bold", size = 12)) +
  
  scale_x_continuous(breaks = round(seq(min(rentalsBYtemp$temperature), max(rentalsBYtemp$temperature), by = 5),1)) +
  scale_colour_manual(values = c("2011" = "#FFA500", "2012" = "#02075d")) +
  
  ggtitle(label = "Figure 2",
          subtitle = "Median number of rentals per day by temperature and year") +
  labs(caption = "Data source: bike.rental.data") +
  ylab("Median number of daily rentals") + xlab("Temperature (°C)")


```

Figure 2 shows that the average number of daily rentals increases with the tempertaure until around 24-26 degrees where it starts decreasing. 

